A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-08 paper.tex

\section*{The Event ARG data structure}\label{eARG}

In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we call it an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

\section*{Node annotations}
[FIXME: this is not the intended transition or next step, just putting this
content here for now.]

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass
alternative patterns of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

% This method for representing recombinations may seem wasteful,
% since more than one node is required for each recombination,
% and genomic interval information must be stored on every edge.
% However, asymptotically the amount of information stored is the
% same, and there is much simplicity to be gained by having
% a single type of node and by treating all edges in the same way.
% More importantly --- as we see in the next section --- we gain the
% ability to simplify the ARG by removing nodes without affecting the
% local tree topologies, and also the ability to record only the
% regions of genome relevant to the final, sample genomes. This last
% feature turns out to be the key to efficient processing of the ARG.

% [Possibly more stuff here: clarify why we assign coordinates to edges
% rather than nodes. Basically lay out all the stuff that we can
% talk about \emph{without} going in to the details of
% ancestry resolution.]

>>>>

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we call it an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

\section*{Node vs edge annotations}
Another aspect of the classical description of an ARG data structure that
derives from its historical origins as a stochastic
process is how we encode information about genetic
inheritance in the graph. The approach taken by Griffiths to
doing this is concise, elegant: all we need to do is store
the location of the breakpoint at each recombination event.
Storing a breakpoint for each recombinatin event is a very
natural and straightforward thing to do when simulating
a (``Big''; see Appendix XXX) ARG process.
Given the graph topology and these breakpoints, we can then
recover the
local trees at every point in the genome
by tracing rootwards from the leaves and choosing the
appropiate outbound path at recombination nodes.
However, there are several significant disadvantages
to this approach: [TODO be concrete here later
and give pointers to the section where we discuss the
alternative.]

Fortunately, there is a straightforward change to the
classical eARG encoding that simplifies
implementations and greatly increases the flexibility
of both the types and granularity of inheritance information
that we can represent.
The main change that we need to make is to
associate the
the details of inheritance with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules.
This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes (see section XXX).
Whether one regards nodes as genomes or events
is something of a philosophical question, and mostly unimportant
from a computational perspective.
